---
title: "Phillip and Robinson's model"
author: "Pablo Rodríguez-Sánchez"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: yes
  md_document:
  html_vignette: default
header-includes:
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usetikzlibrary{shapes,arrows}
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(sleepR)
```

# Model
The function `philrob` simulates Phillips and Robinson's model for sleep-wake dynamics.

## Dynamics
The dynamics of the Phillips and Robinson model are given by the following system of ordinary differential equations:

$$\left[
\begin{array}{c}
\tau_v \dot V_v + V_v \\
\tau_m \dot V_m + V_m \\
\chi \dot H + H
\end{array}
\right]
=
\left[
\begin{array}{ccc}
0 & -\nu_{vm} & \nu_{vh} \\
-\nu_{mv} & 0 & 0 \\
0 & \mu & 0
\end{array}
\right]
\left[
\begin{array}{c}
S(V_v) \\
S(V_m) \\
H
\end{array}
\right]
+
\left[
\begin{array}{c}
-\nu_{vc} C(t) \\
\nu_{ma} S(V_{a0}) \\
0
\end{array}
\right]$$

where $S(V)$ is the saturation function:

$$S(V) = \frac{Q_{max}}{1 + e^{-\frac{V-\theta}{\sigma}}}$$

```{r, Plot-saturation, include = TRUE, echo = FALSE, fig.height = 3}
parms <- philrob_default_parms()
S <- function(V) {
  parms['Qmax']/(1 + exp((parms['theta']-V)/parms['sigma']))
}

xs <- seq(-10, 35, length.out = 100)
plot(xs, S(xs), type = 'l', xlab = 'V', ylab = 'S(V)', col = 'red', axes = FALSE)

# Asymptote
lines(c(-10, 35), c(parms['Qmax'], parms['Qmax']), lty = 'dashed')

# Half life
lines(c(-10, parms['theta']), c(parms['Qmax']/2, parms['Qmax']/2), lty = 'dashed')
lines(c(parms['theta'], parms['theta']), c(0, parms['Qmax']/2), lty = 'dashed')

axis(2, at = c(0, parms['Qmax']/2, parms['Qmax']), labels = c('0', 'Qmax/2', 'Qmax'))
axis(1, at = c(-10, 0, parms['theta'], 35), labels = c('', '0', 'theta', ''))
```

and the external forcing is typically given by $C(t)$:

$$C(t) = \frac{1}{2} \left( 1 + \cos(\omega t + \alpha) \right)$$

\newpage

## Parameters

The default values for the parameters are listed in the table below:

| Symbol              |      Value       | Units              |
|:--------------------|:----------------:|:-------------------|
| $\tau_m$            |    $10/3600$     | $h$                |
| $\tau_v$            |    $10/3600$     | $h$                |
| $\chi$              |      $10.8$      | $h$                |
| $\nu_{vm}$          |    $1.9/3600$    | $mV \cdot h$       |
| $\nu_{mv}$          |    $1.9/3600$    | $mV \cdot h$       |
| $\nu_{vh}$          |      $0.19$      | $mV \cdot nM^{-1}$ |
| $\mu$               |    $10^{-3}$     | $nM \cdot h$       |
| $\nu_{vc}$          |      $6.3$       | $mV$               |
| $\nu_{ma}S(V_{a0})$ |       $1$        | $mV$               |
| $Q_{max}$           | $100 \cdot 3600$ | $h^{-1}$           |
| $\theta$            |       $10$       | $mV$               |
| $\sigma$            |       $3$        | $mV$               |
| $\omega$            |   $2 \pi / 24$   | $h^{-1}$           |
| $\alpha$            |       $0$        | $1$                |

## State variables

The state variables are defined as:

| State variable | Units | Physiological interpetation | Informal interpretation |
|:---------------|:-----:|:----------------------------|:------------------------|
| $V_v$          | $mV$  | Activity of the VLPO        | Stay asleep system      |
| $V_m$          | $mV$  | Activity of the MA          | Stay awake system       |
| $H$            |  $1$  | Homeostatic pressure        | Somnogen level          |

## Diagram

\begin{center}
\begin{tikzpicture}[node distance = 2cm, auto]

    % Definitions
    \tikzstyle{state} = [rectangle,draw,fill=blue!20,text width=5em,text centered,rounded corners,minimum height=4em]
    \tikzstyle{line} = [draw, -latex']
    \tikzstyle{source} = [draw, ellipse,fill=red!20, node distance=3cm,minimum height=2em]

    % Place nodes
    % States
    \node [state] (Vv) {$V_v$};
    \node [state, below of=Vv] (Vm) {$V_m$};
    \node [state, below of=Vm] (H) {$H$};

    % Sources
    \node [source, right of=Vv] (C) {$C(t)$};
    \node [source, right of=Vm] (A) {$A$};

    % Draw edges
    % Sources
    \path [blue, line] (C) -> node{$\nu_{vc}$} (Vv);
    \path [line] (A) -> node{$\nu_{ma}$} (Vm);

    % Coupling
    \path [line] (Vm) -> node{$\mu$} (H);
    \draw[red, ->] (Vv) -- node[midway, right] {$\nu_{mv}$} (Vm);
    \draw[red, ->] (Vm) -- node[midway, left] {$\nu_{vm}$} (Vv);
    \draw[->] (H.west) .. controls +(left:20mm) and +(left:20mm) .. node{$\nu_{vh}$} (Vv.west);

    % Decay
    \draw[red, ->] (Vv.north) .. controls +(up:10mm) and +(up:10mm) .. (Vv.north west);
    \draw[red, ->] (Vm.north west) .. controls +(left:10mm) and +(left:10mm) .. (Vm.south west);
    \draw[red, ->] (H.south) .. controls +(down:10mm) and +(down:10mm) .. (H.south west);

\end{tikzpicture}
\end{center}

Schematic summary of the dynamics. The blue nodes represent the system's states ($V_v$ the activity of the ventrolateral preoptic area, $V_m$ the activity of the mono aminergic group and $H$ the homeostatic pressure). The red nodes represent the external sources ($C(t)$, the astronomical light/dark forcing, and $A$, the acetylcholine group constant influence). The positive effects are coded as black arrows. Negative ones as red arrows. Blue arrows represent oscillating effects.

## Reference

Phillips AJK, Robinson PA.
A Quantitative Model of Sleep-Wake Dynamics Based on the Physiology of the Brainstem Ascending Arousal System.
J Biol Rhythms. 2007 Apr 29;22(2):167–79. Available from: http://journals.sagepub.com/doi/10.1177/0748730406297512

\newpage

# Examples of usage

## Getting the time series

With default parameters:

```{r Problem posing with default pars, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

# Simulate
sol <- philrob(ts, y0)

```

With custom parameters:

```{r Problem posing with overriden pars, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 3
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms() # Load default parameters...
parms['vvc'] <- 6 # .. and modify one

# Simulate
sol <- philrob(ts, y0, parms)
```

With custom forcing:

```{r Problem posing with custon forcing, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 3
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

C <- function(t) { 0 }

# Simulate
sol <- philrob(ts, y0, parms = philrob_default_parms(), forcing = C)
```

With stabilization run of three days:

```{r Problem posing with stabilization run, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

# Simulate
sol <- philrob(ts, y0, tStabil = 3*24)

```

The output looks like:

```{r Display output, echo=FALSE, results='asis'}
knitr::kable(head(sol, 5))
```

where:

- `time`: the time (in h),
- `Vv`: activity of the ventrolateral preoptic area (in mV)
- `Vm`: activity of the monoaminergic group (in mV)
- `H`: homeostatic pressure / somnogen
- `asleep`: the asleep/awake status (`TRUE` if asleep, `FALSE` if awake)

## Plotting results

### Raster / somnogram plot

```{r Plotting, include = TRUE, fig.show='hold'}
philrobPlot(sol)
rasterPlot(sol)
```
