---
title: "Phillip and Robinson's model"
author: "Pablo Rodríguez-Sánchez"
date: "`r Sys.Date()`"
output: 
  pdf_document:
  html_vignette: default
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usetikzlibrary{shapes,arrows}
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(sleepR)
```

# Model
The function `philrob` simulates Phillips and Robinson's model for sleep-wake dynamics.

## Dynamics
The dynamics of the Phillips and Robinson model are given by the following system of ordinary differential equations:

$$\left[
\begin{array}{c}
\tau_v \dot V_v + V_v \\
\tau_m \dot V_m + V_m \\
\chi \dot H + H
\end{array}
\right]
=
\left[
\begin{array}{ccc}
0 & -\nu_{vm} & \nu_{vh} \\
-\nu_{mv} & 0 & 0 \\
0 & \mu & 0
\end{array}
\right]
\left[
\begin{array}{c}
S(V_v) \\
S(V_m) \\
H
\end{array}
\right]
+
\left[
\begin{array}{c}
-\nu_{vc} C(t) \\
\nu_{ma} S(V_{a0}) \\
0
\end{array}
\right]$$

where $S(V)$ is the saturation function:

$$S(V) = \frac{Q_{max}}{1 + e^{-\frac{V-\theta}{\sigma}}}$$

and the external forcing is given by $C(t)$:

$$C(t) = \frac{1}{2} \left( 1 + cos(\omega t + \alpha) \right)$$

## State variables

The state variables are defined as:

| State variable | Physiological interpetation | Informal interpretation |
|:---------------|:----------------------------|:------------------------|
| $V_v$          | Activity of the VLPO        | Stay asleep system      |
| $V_m$          | Activity of the MA          | Stay awake system       |
| $H$            | Homeostatic pressure        | Somnogen level          |

## Diagram

\begin{center}
\begin{tikzpicture}[node distance = 2cm, auto]

    % Definitions
    \tikzstyle{state} = [rectangle,draw,fill=blue!20,text width=5em,text centered,rounded corners,minimum height=4em]
    \tikzstyle{line} = [draw, -latex']
    \tikzstyle{source} = [draw, ellipse,fill=red!20, node distance=3cm,minimum height=2em]
    
    % Place nodes
    % States
    \node [state] (Vv) {$V_v$};
    \node [state, below of=Vv] (Vm) {$V_m$};
    \node [state, below of=Vm] (H) {$H$};
    
    % Sources
    \node [source, right of=Vv] (C) {$C(t)$};
    \node [source, right of=Vm] (A) {$A$};
    
    % Draw edges
    % Sources
    \path [blue, line] (C) -> node{$\nu_{vc}$} (Vv);
    \path [line] (A) -> node{$\nu_{ma}$} (Vm);
    
    % Coupling
    \path [line] (Vm) -> node{$\mu$} (H);
    \draw[red, ->] (Vv) -- node[midway, right] {$\nu_{mv}$} (Vm);
    \draw[red, ->] (Vm) -- node[midway, left] {$\nu_{vm}$} (Vv);
    \draw[->] (H.west) .. controls +(left:20mm) and +(left:20mm) .. node{$\nu_{vh}$} (Vv.west);
    
    % Decay
    \draw[red, ->] (Vv.north) .. controls +(up:10mm) and +(up:10mm) .. (Vv.north west);
    \draw[red, ->] (Vm.north west) .. controls +(left:10mm) and +(left:10mm) .. (Vm.south west);
    \draw[red, ->] (H.south) .. controls +(down:10mm) and +(down:10mm) .. (H.south west);

\end{tikzpicture}
\end{center}

Schematic summary of the dynamics. The blue nodes represent the system's states ($V_v$ the activity of the ventrolateral preoptic area, $V_m$ the activity of the mono aminergic group and $H$ the homeostatic pressure). The red nodes represent the external sources ($C(t)$, the astronomical light/dark forcing, and $A$, the acetylcholine group constant influence). The positive effects are coded as black arrows. Negative ones as red arrows. Blue arrows represent oscillating effects.

## Reference

Phillips AJK, Robinson PA.
A Quantitative Model of Sleep-Wake Dynamics Based on the Physiology of the Brainstem Ascending Arousal System.
J Biol Rhythms. 2007 Apr 29;22(2):167–79. Available from: http://journals.sagepub.com/doi/10.1177/0748730406297512

\newpage

# Examples of usage

## Getting the time series

With default parameters:

```{r Problem posing with default pars, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

# Simulate
sol <- philrob(ts, y0)

```

With custom parameters:

```{r Problem posing with overriden pars, include = TRUE}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 3
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms() # Load default parameters...
parms['vvc'] <- 6 # .. and modify one

# Simulate
sol <- philrob(ts, y0, parms)
```

The output looks like:

```{r Display output, echo=FALSE, results='asis'}
knitr::kable(head(sol, 5))
```

where:

- `time`: the time (in h),
- `Vv`: activity of the ventrolateral preoptic area (in mV)
- `Vm`: activity of the monoaminergic group (in mV)
- `H`: homeostatic pressure / somnogen
- `asleep`: the asleep/awake status (`TRUE` if asleep, `FALSE` if awake)

## Plotting results

### Raster / somnogram plot

```{r Plotting, include = TRUE, fig.show='hold'}
philrobPlot(sol)
rasterPlot(sol)
```

\newpage

# Results

In this section, we analyze the effect of removing (equating to zero) one by one the main elements in the defining equation of the model. We'll refer to them as:

- `Coupling`: the elements in the square matrix.
- `Source`: the constant term $\nu_{ma} S(V_a)$.
- `Forcing`: the time dependent term $-\nu_{vc} C(t)$.

## Case 000

### Settings
- `Coupling`: no
- `Source`: no
- `Forcing`: no

### Results
All states decay to zero.

```{r 000, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Coupling
parms['vmv'] <- 0
parms['vvm'] <- 0
parms['vvh'] <- 0
parms['mu'] <- 0

# Source
parms['vmaSa'] <- 0

# Forcing
parms['vvc'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```

## Case 001

### Settings
- `Coupling`: no
- `Source`: no
- `Forcing`: yes

### Results
All states decay to zero, with the exception of $V_v$, that keeps oscillating in synchrony with the forcing.

```{r 001, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Coupling
parms['vmv'] <- 0
parms['vvm'] <- 0
parms['vvh'] <- 0
parms['mu'] <- 0

# Source
parms['vmaSa'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```


## Case 010

### Settings
- `Coupling`: no
- `Source`: yes
- `Forcing`: no

### Results
All states decay to constant values. $V_v$ and $H$ decay to zero, and $V_m$ to a non-zero constant value.

```{r 010, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Coupling
parms['vmv'] <- 0
parms['vvm'] <- 0
parms['vvh'] <- 0
parms['mu'] <- 0

# Forcing
parms['vvc'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```


## Case 011

### Settings
- `Coupling`: no
- `Source`: yes
- `Forcing`: yes

### Results
All states decay to constant values, with the exception of $V_v$, that keeps oscillating in synchrony with the forcing.

```{r 011, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Coupling
parms['vmv'] <- 0
parms['vvm'] <- 0
parms['vvh'] <- 0
parms['mu'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```


## Case 100

### Settings
- `Coupling`: yes
- `Source`: no
- `Forcing`: no

### Results
All states decay to constant values. 

```{r 100, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Source
parms['vmaSa'] <- 0

# Forcing
parms['vvc'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```


## Case 101

### Settings
- `Coupling`: yes
- `Source`: no
- `Forcing`: yes

### Results
A very reasonable behaviour arises, despite the absence of the source term.

```{r 101, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Source
parms['vmaSa'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```

## Case 110

### Settings
- `Coupling`: yes
- `Source`: yes
- `Forcing`: no

### Results
A periodic behavior appears despite the absence of the forcing term. Like this, the system is autonomous. The period of the oscillations is around 8 hours.

```{r 110, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Forcing
parms['vvc'] <- 0

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```

## Case 111

### Settings
- `Coupling`: yes
- `Source`: yes
- `Forcing`: yes

### Results
This is the result described in the abovementioned paper.

```{r 111, include = TRUE, echo = FALSE, fig.show='hold'}
## Problem setting
y0 <- c(Vv = -13, Vm = 1, H = 10) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

parms <- philrob_default_parms()

# Simulate
sol <- philrob(ts, y0, parms)

# Plot
philrobPlot(sol)
rasterPlot(sol)
```
