---
title: "Strogatz's model"
author: "Pablo Rodríguez-Sánchez"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: yes
  html_vignette: default
vignette: >
  %\VignetteIndexEntry{Strogatz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(sleepR)
```
# Model
The function `strogatz` simulates Strogatz's model for sleep-wake dynamics. The dynamics have the following structure:

$$
\dot \theta_1 = \omega_1 - C_1 \cdot \cos(2 \pi (\theta_2 - \theta_1)) \\
\dot \theta_2 = \omega_2 + C_2 \cdot \cos(2 \pi (\theta_2 - \theta_1))
$$

# Analysis

It is useful to note that, using the relative phase $\psi \equiv \theta_1 - \theta_2$, the system can be simplified to:

$$
\dot \psi = \Omega - C \cos(2 \pi \psi)
$$
where $\Omega \equiv \omega_1 - \omega_2$ and $C \equiv C_1 + C_2$. The relative phase can be stabilized if and only if $\dot \psi = 0$, and this is only possible if:

$$
\lvert \frac{\Omega}{C} \rvert \leq 1
$$

# Reference

Strogatz, S. H. (1987).
Human sleep and circadian rhythms: a simple model based on two coupled oscillators.
Journal of Mathematical Biology, 25(3), 327–347. http://doi.org/10.1007/BF00276440

\newpage

# Examples of usage

## Getting the time series

With default parameters:

```{r Problem posing with default pars, include = TRUE}
## Problem setting
y0 <- c(th1 = 0.1, th2 = 0.05) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

# Simulate
sol <- strogatz(ts, y0)

```

With custom parameters:

```{r Problem posing with overriden pars, include = TRUE}
## Problem setting
y0 <- c(th1 = 0.1, th2 = 0.05) # Initial conditions

nDays <- 5
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate
parms <- c(w1=1/24, w2=0.85/24, C1=0/24, C2=0.16/24) # Parameters

# Simulate
sol <- strogatz(ts, y0, parms)
```

The output looks like:

```{r Display output, echo=FALSE, results='asis'}
knitr::kable(head(sol, 5))
```

where:

- `time`: the time (in h), 
- `th1`: the phase of the circadian oscillator
- `th2`: the phase of the sleep-wake cycle
- `asleep`: the asleep/awake status (`TRUE` if asleep, `FALSE` if awake)

## Plotting results

### Raster / somnogram plot

```{r Plotting, include = TRUE, warning = FALSE, fig.show='hold'}
plot(sol$time/24, sol$asleep, type = 'line', xlab = 'Time (d)', ylab = 'Asleep')
rasterPlot(sol)
```

# Results

## Entrained vs. non entrained case
```{r Entrained vs not, echo = FALSE, include = TRUE, fig.show='hold'}
## Problem setting
y0 <- c(th1 = 0.1, th2 = 0.05) # Initial conditions

nDays <- 60
ts <- seq(0, nDays*24, length.out=nDays*24*20) # Times to simulate

# Parameters
# With this settings, the bifurcation happens at w2 = 0.84/24 h^-1
parms_entrained <- c(w1=1/24, w2=0.845/24, C1=0/24, C2=0.16/24)
parms_non_entrained <- c(w1=1/24, w2=0.835/24, C1=0/24, C2=0.16/24)

# Simulate
sol_entrained <- strogatz(ts, y0, parms_entrained)
sol_non_entrained <- strogatz(ts, y0, parms_non_entrained)

# Plot
rasterPlot(sol_entrained)
title('Entrained')
rasterPlot(sol_non_entrained)
title('Not entrained')
```
